Analysis-ready VCF at Biobank scale using Zarr

### 2024-11-15 paper.tex
Add Will Tyler to authorship list for preprint update

%%% Use the \authfn to add symbols for additional footnotes, if any. 
% 1 is reserved for correspondence emails; then continuing with 2 etc for contributions.
% First author
\author[1,\authfn{1}]{Eric Czech} % https://orcid.org/0000-0002-4254-4255
\author[2,3\authfn{1}]{Timothy R. Millar} % https://orcid.org/0000-0002-5142-8811
\author[4,\authfn{1}]{Tom White}

% Middle
\author[5]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
\author[6]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
\author[7]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
\author[1]{Rafal Wojdyla} % https://orcid.org/0009-0005-0735-7090
\author[8]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284

% Senior 
\author[1,\authfn{2}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
\author[5,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253

\affil[1]{Related Sciences}
\affil[2]{The New Zealand Institute for Plant \& Food Research Ltd, Lincoln,
New Zealand}
\affil[3]{Department of Biochemistry, School of Biomedical Sciences, University of Otago, Dunedin, New Zealand}
\affil[4]{Tom White Consulting Ltd.}
\affil[5]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
University of Oxford, UK}
\affil[6]{Wellcome Sanger Institute}
\affil[7]{Genomics England}
\affil[8]{School of Computer Science, McGill University, Montreal, QC, Canada}

@@

\subsection{Case study: 1,063 spruce whole-genome samples}

To demonstrate the versatility of \texttt{vcf2zarr}, in this section
we include a case study of a dataset originating from a species whose
genome properties differ substantially from human. The conifer Norway
spruce, \emph{Picea abies}, is one of the largest and most
ecologically important species on Earth and is distributed over large
parts of the Northern hemisphere. One of its features is its large
genome, consisting of 12 autosomes with a genome size of 19.6
Gb~\cite{nystedt_NorwaySpruceGenome_2013}. Here, we assess the
performance of \texttt{vcf2zarr} as applied to 1,063 whole-genome
resequenced spruce individuals from a recent study (in preparation),
where resequencing data has been mapped to a chromosome-scale
reference genome. The dataset consists of approximately 3,745 million
single-nucleotide variants and small indels from the 12 autosomes
totalling 7.4TiB of VCF data after \texttt{bgzip} compression,
distributed over 165 VCF files. Since the dataset originally was
generated for downstream analyses that require genotype likelihoods,
the genotypes include the PL field.

%% Regarding data storage files, which to use?
%% find spruce.icf -name "*" | wc -l -> 1394080
%% # chunks in spruce.icf.inspect.txt -> 1342569
%% du -hs spruce.icf -> 6.8TiB
%% sum size in spruce.icf.inspect.txt -> 6.8TiB
%% du -hs spruce.vcz -> 6.7TiB
%% sum size in spruce.vcz.inspect.txt -> 6.6TiB
%% # chunks in spruce.vcz.inspect.txt -> 11984581
%% More chunk files despite smaller input data: could selection of
%% chunk size matter here?
Following the conversion process detailed in the previous section, we
first converted the 165 VCF files (7.33TiB) to ICF, followed by
conversion to Zarr. The tasks were run as single jobs on a compute
node with 2,048 GB RAM and 128 cores. Conversion to ICF required
13h45min runtime. The ICF representation used a total of 6.77 TiB over
1,342,569 data storage files. Conversion to Zarr required 18h51min
runtime, generating a dataset with 32 arrays, consuming a total of 6.6
TiB storage over 11,984,581 chunk files. The reduction in total
storage (1.1X) is negligible, mainly due to the inherent difficulty in
compressing the PL field. The top fields in terms of storage are
detailed in Table~\ref{tab-spruce-data}.

\begin{table}
\caption{Summary for a selection of the largest VCF Zarr columns produced for 
1,063 VCFs on 12 spruce chromosomes using \texttt{vcf2zarr}
default settings. For column details see the caption of Table~\ref{tab-genomics-england-data}.
\label{tab-spruce-data}}
\begin{tabular}{llS[table-format=3.1]S[table-format=3.2]S[table-format=3.2]}
\toprule
{Field} & {type} & {storage} & {compress} & {\%total} \\
  \midrule
  /call\_PL & int16 & 6.12 TiB & 12.0 & 92.75\% \\
  /call\_genotype & int8 & 282.45 GiB & 26.0 & 4.18\% \\
  /call\_genotype\_mask & bool & 42.15 GiB & 180.0 & 0.62\% \\
  /variant\_DP4 & int32 & 26.24 GiB & 2.1 & 0.39\% \\
  /variant\_MQSBZ & float32 & 13.08 GiB & 1.1 & 0.19\% \\
  /variant\_RPBZ & float32 & 13.04 GiB & 1.1 & 0.19\% \\
  /variant\_MQBZ & float32 & 12.97 GiB & 1.1 & 0.19\% \\
  /variant\_SCBZ & float32 & 12.94 GiB & 1.1 & 0.19\% \\
  /variant\_VDB & float32 & 12.84 GiB & 1.1 & 0.19\% \\
  /variant\_quality & float32 & 12.84 GiB & 1.1 & 0.19\% \\
  /variant\_BQBZ & float32 & 12.64 GiB & 1.1 & 0.19\% \\
  /variant\_SGB & float32 & 12.12 GiB & 1.2 & 0.18\% \\
  /variant\_position & int32 & 9.65 GiB & 1.4 & 0.14\% \\
  /variant\_AC & int16 & 7.42 GiB & 2.8 & 0.11\% \\
  /variant\_DP & int32 & 6.71 GiB & 2.1 & 0.10\% \\
  /variant\_allele & object & 5.37 GiB & 21.0 & 0.08\% \\
  /variant\_AN & int16 & 3.05 GiB & 2.3 & 0.05\% \\
  /call\_genotype\_phased & bool & 2.03 GiB & 1800.0 & 0.03\% \\
  /variant\_filter & bool & 1.46 GiB & 2.4 & 0.02\% \\
  /variant\_MQ & int8 & 489.83 MiB & 7.3 & 0.01\% \\
  \bottomrule
\end{tabular}
\end{table}

Table~\ref{tab-spruce-data} shows that the dataset storage size is
dominated by a major component, call\_PL, which accounts for nearly
93\% of the total. call\_PL is yet another field with low compression potential due to inherent noisiness.

%% FIXME: reference regarding VCF POS?
The spruce data set poses an interesting challenge to the VCF storage
format due to the large chromosome sizes (>1Gbp). The VCF POS field is
encoded in 32-bit format (corresponding to a maximum of approximately
537Mbp), which mandates splitting up larger chromosomes into chunks,
adding additional complexity to downstream analyses. Zarr does not
impose this limitation, and we therefore demonstrate the versatility
of Zarr by updating contig information and the coordinate system for
the spruce dataset.

@@

\section{Acknowledgements}
We would like to thank Gil McVean for helpful comments on the manuscript.

This research was made possible through access to data in the National Genomic
Research Library, which is managed by Genomics England Limited (a wholly owned
company of the Department of Health and Social Care). The National Genomic
Research Library holds data provided by patients and collected by the NHS as
part of their care and data collected as part of their participation in
research. The National Genomic Research Library is funded by the National
Institute for Health Research and NHS England. The Wellcome Trust, Cancer
Research UK and the Medical Research Council have also funded research
infrastructure.

>>>>

%%% Use the \authfn to add symbols for additional footnotes, if any. 
% 1 is reserved for correspondence emails; then continuing with 2 etc for contributions.
% First author
\author[1,\authfn{1}]{Eric Czech} % https://orcid.org/0000-0002-4254-4255
\author[2,3\authfn{1}]{Timothy R. Millar} % https://orcid.org/0000-0002-5142-8811
\author[4,\authfn{1}]{Will Tyler}
\author[5,\authfn{1}]{Tom White}

% Middle
\author[6]{Ben Jeffery} % https://orcid.org/0000-0002-1982-6801
\author[7]{Alistair Miles} % https://orcid.org/0000-0001-9018-4680
\author[8]{Sam Tallman} % https://orcid.org/0000-0001-7183-6276
\author[1]{Rafal Wojdyla} % https://orcid.org/0009-0005-0735-7090
\author[9]{Shadi Zabad} % https://orcid.org/0000-0002-8003-9284

% Senior 
\author[1,\authfn{2}]{Jeff Hammerbacher} % https://orcid.org/0000-0001-6596-8563
\author[6,\authfn{2},\authfn{3}]{Jerome Kelleher} % https://orcid.org/0000-0002-7894-5253

\affil[1]{Related Sciences}
\affil[2]{The New Zealand Institute for Plant \& Food Research Ltd, Lincoln,
New Zealand}
\affil[3]{Department of Biochemistry, School of Biomedical Sciences, University of Otago, Dunedin, New Zealand}
\affil[4]{Independent researcher}
\affil[5]{Tom White Consulting Ltd.}
\affil[6]{Big Data Institute, Li Ka Shing Centre for Health Information and Discovery, 
University of Oxford, UK}
\affil[7]{Wellcome Sanger Institute}
\affil[8]{Genomics England}
\affil[9]{School of Computer Science, McGill University, Montreal, QC, Canada}

@@

@@

\section{Acknowledgements}
% We would like to thank Gil McVean for helpful comments on the manuscript.
This research was made possible through access to data in the National Genomic
Research Library, which is managed by Genomics England Limited (a wholly owned
company of the Department of Health and Social Care). The National Genomic
Research Library holds data provided by patients and collected by the NHS as
part of their care and data collected as part of their participation in
research. The National Genomic Research Library is funded by the National
Institute for Health Research and NHS England. The Wellcome Trust, Cancer
Research UK and the Medical Research Council have also funded research
infrastructure.
